---
title: 'Nashville Crime Web App Part 2: Interactive Visuals'
author: Coy McNew
date: '2020-10-05'
slug: nashville-crime-web-app-part-2-interactive-visuals
categories:
  - Nashville Crime Web App
tags:
  - leaflet
  - sf
  - geospatial
# draft: true
---

# Objective

The objective of this series of posts is to catalog my effort on a project which accomplishes the following:

1. Connect directly to publicly available data from data.nashville.gov
2. Summarize the information in an interactive map and time series
3. Mine text description fields to summarize in a visual and allow for keyword searching
4. Run as a Shiny web app, hosted on [shinyapps.io](shinyapps.io) for anyone to use

In this, the second post of a four part series, I'm discussing one of the interactive visuals.  We're going to be making an interactive map using `leaflet` to sptially display the crime data we pulled in [Part 1](https://www.coymcnew.com/post/2020/09/08/nashville-crime-web-app-part-1-getting-the-data/).

# Mapping

There are several packages to develop very cool and polished interactive maps in R.  My favorites are [`leaflet`](https://github.com/rstudio/leaflet), [`mapview`](https://github.com/r-spatial/mapview) and [`tmap`](https://github.com/mtennekes/tmap).  All three can be used to output html widgets and can be integrated into shiny web apps.  `mapview` makes a really excellent looking default map for quick viewing of a spatial dataset, but I tend to prefer `leaflet` for its customizability and my familiarity.

## Geospatial Data in R

There has been a lot written about working with geospatial data in R.  My favorite resource is a fairly comprehensive guide by Robin Lovelace, Jakub Nowosad, and Jannes Muenchow called [Geocomputation with R](https://geocompr.robinlovelace.net).  There are several packages that make working with geospatial data more manageable in R, most notably the [`sf`](https://github.com/r-spatial/sf) package.  `sf` allows for the grouping of coordinates into a special geometry column on the right side of a dataframe, turning it into an object of class "sf".  This geometry column is sticky, meaning when you use typical tidy functions like filter or select, the geometry data remains attached to each corresponding row.  Many mapping packages recognize these "sf" objects as geospatial objects and so they use the appropriate geometry data automatically without explicitly assigning.   

## `leaflet` 

Below I'll load the 500 most recent observations from the Nashville crime database, as described in [Part 1](https://www.coymcnew.com/post/2020/09/08/nashville-crime-web-app-part-1-getting-the-data/) and convert the dataframe to an "sf" object.

``` {r message=FALSE, warning=FALSE}
library(tidyverse)
library(RSocrata)
library(sf)
#load 500 most recent observations
mapDF <- read.socrata("https://data.nashville.gov/resource/2u6v-ujjs.csv?$order=incident_occurred DESC &$limit=500") %>%
  #drop rows with empty coordinates
  filter(!is.na(latitude) | !is.na(longitude)) 
#convert to sf
mapSF <- mapDF %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
#take a quick look
str(mapSF)
```

Then to get a quick look, we'll pass mapSF to leaflet, add some default basemap tiles, and add rows as circle markers, with some minor alterations to the defulat color and size.

```{r message=FALSE, warning=FALSE}
library(leaflet)
leaflet(mapSF) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 3, color = 'black', opacity = 1, 
    fillColor = 'blue', fillOpacity = 0.4, weight = 0.5
  )
```

### Basemap

That looks pretty good, but let's improve some default components and add some functionality and polish.  First, let's change the default basemap and add a basemap selection option.  `leaflet` provides a multitude of basemap options, and you can preview the options from a handy [web app](https://leaflet-extras.github.io/leaflet-providers/preview/).  Since I like a very unobtrusive default basemap, I'll add CartoDB.Positron as the default option, with the popular OpenStreetMap.HOT as a toggle option for those who want more information.  Basemaps are added via `addProviderTiles()` and the basemap names are provided in the list `providers`.  I'll also add these as toggleable base layers by assigning a group name and then adding the groups as `baseGroups` in the `addLayersControl()` call.  

```{r message=FALSE, warning=FALSE}
leaflet(mapSF) %>%
  addProviderTiles(providers$CartoDB.Positron, group = 'CartoDB') %>%
  addProviderTiles(providers$OpenStreetMap.HOT, group = 'Open Street Map') %>%
  addCircleMarkers(
    radius = 3, color = 'black', opacity = 1, 
    fillColor = 'blue', fillOpacity = 0.4, weight = 0.5
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "Open Street Map"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Tooltips

Ok, the basemap is set so now let's add some functionality and polish with some tooltip information on mouse hover.  The `label` argument within `addCircleMarkers()` allows us to add text information as html.  I'm going to use the [`htmltools`](https://github.com/rstudio/htmltools) package to help me apply some formatting and line breaks here.  You can just as easily type out the html needed, but I like to use the helper functions. I'm also going to make sure R understands we're supplying html by using `lapply(..., HTML)`.  As for what to include, I want to see a description of the crime and date of occurrence, address, and victim information so I'm going to pull the relevant information from each corresponding column.

```{r message=FALSE, warning=FALSE}
library(htmltools)
leaflet(mapSF) %>%
  addProviderTiles(providers$CartoDB.Positron, group = 'CartoDB') %>%
  addProviderTiles(providers$OpenStreetMap.HOT, group = 'Open Street Map') %>%
  addCircleMarkers(
    radius = 3, color = 'black', opacity = 1, 
    fillColor = 'blue', fillOpacity = 0.4, weight = 0.5,
    label = ~lapply(paste0(
      tags$b("Description"), ": ", offense_description,
      br(), tags$b("Date"), ": ", incident_occurred,
      hr(),
      tags$b("Address"), ": ", incident_location, " ", zip_code,
      hr(),
      tags$b("Victim Information"),
      br(), "Gender: ", victim_gender, ", Race: ", victim_race, ", Ethnicity: ", victim_ethnicity
    ), HTML)
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "Open Street Map"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Color

Now to step this map up a notch, lets add some color coding for one of the variables.  My original thought was to color code by date so I could see some sort of pattern to the time series, but the result looked pretty messy and I saw no pattern, as can be seen in the header of [Part 1](https://www.coymcnew.com/post/2020/09/08/nashville-crime-web-app-part-1-getting-the-data/).  Another interesting variable to me is victim_county_resident because I want to see how much crime is happening to tourists vs locals.  I expect to see far more tourism related crime downtown.  First, I'm going to clean up this variable by changing the values to "Yes", "No", or "Unknown" and make this variable an ordered factor for better display.

```{r message=FALSE, warning=FALSE}
mapSF <- mapSF %>%
  mutate(county_resident_YN = 
           case_when(
              victim_county_resident == "RESIDENT"         ~  "Yes",
              victim_county_resident == "NON RESIDENT"     ~  "No",
              victim_county_resident %in% c("", "UNKNOWN") ~  "Unknown"
           ) %>%
           factor(levels = c("Yes", "No", "Unknown"), ordered = TRUE)
  )
```

`leaflet` provides several helper functions to define color mappings functions for use within map building.  An explanation of these functions can be found [here](https://rstudio.github.io/leaflet/colors.html).  Below, I'm going to use `colorFactor()` to map the values of our new residence variable and call this color function `cols()`.  I'll then define our fill colors by passing our new residence variable to the `cols()` function.

```{r message=FALSE, warning=FALSE}
cols <- colorFactor("viridis", mapSF$county_resident_YN)
leaflet(mapSF) %>%
  addProviderTiles(providers$CartoDB.Positron, group = 'CartoDB') %>%
  addProviderTiles(providers$OpenStreetMap.HOT, group = 'Open Street Map') %>%
  addCircleMarkers(
    radius = 3, color = 'black', opacity = 1, 
    fillColor = ~cols(county_resident_YN), fillOpacity = 0.6, weight = 0.5,
    label = ~lapply(paste0(
      tags$b("Description"), ": ", offense_description,
      br(), tags$b("Date"), ": ", incident_occurred,
      hr(),
      tags$b("Address"), ": ", incident_location, " ", zip_code,
      hr(),
      tags$b("Victim Information"),
      br(), "Gender: ", victim_gender, ", Race: ", victim_race, ", Ethnicity: ", victim_ethnicity
    ), HTML)
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "Open Street Map"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Legend

Ok now to cap off this map visual, let's add a color legend for our county residence variable.  `leaflet` makes this easy since we defined our color mapping function up front and since we took the time to define our variable as an ordered facotr, the mapped values will show up in the correct order on the legend.  To define the legend, we simply use the `addLegend()` call and pass our color function and variable used for coloring.

```{r message=FALSE, warning=FALSE}
leaflet(mapSF) %>%
  addProviderTiles(providers$CartoDB.Positron, group = 'CartoDB') %>%
  addProviderTiles(providers$OpenStreetMap.HOT, group = 'Open Street Map') %>%
  addCircleMarkers(
    radius = 3, color = 'black', opacity = 1, 
    fillColor = ~cols(county_resident_YN), fillOpacity = 0.6, weight = 0.5,
    label = ~lapply(paste0(
      tags$b("Description"), ": ", offense_description,
      br(), tags$b("Date"), ": ", incident_occurred,
      hr(),
      tags$b("Address"), ": ", incident_location, " ", zip_code,
      hr(),
      tags$b("Victim Information"),
      br(), "Gender: ", victim_gender, ", Race: ", victim_race, ", Ethnicity: ", victim_ethnicity
    ), HTML)
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "Open Street Map"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend("bottomright", pal = cols, values = ~county_resident_YN,
    title = lapply(paste0("Was the victim", br(), "a resident of", br(), "Davidson County?"), HTML),
    opacity = 1
  )
```

# Conclusions

Alright, I think that map looks pretty good!  It sums up all of the information we care about and is visually pleasing so I think we'll move forward with it.  Since the map took a bit longer than I thought it would, I'll save the other visual for another post, Part 2.5 of the series...

In conclusion we accomplished the following in this post:

1. Generated an interactive map of the crime incident data we connected to in [Part 1](https://www.coymcnew.com/post/2020/09/08/nashville-crime-web-app-part-1-getting-the-data/).
2. Added a couple basemap options on toggle.
3. Added crime incident information on mouse hover.
4. Created a new variable of victim county residence and color mapped it.

The next step is to look at the time series component of the data and generate an interesting visual to summarize it.


